Two subtle problems with over-representation analysis

### 2024-06-11 manuscript/manuscript.Rmd
mswork

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 500 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected algorithm for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets over seven independent datasets detected with clusterProfiler's
`enricher` function with and without correcting the background problem.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

@@

A workaround for the FDR problem was developed and as above, and original and corrected
analyses were conducted for the seven datasets and nine gene set libraries.
The results obtained showed no major differences in the number of significant sets between
original and corrected analyses (**Figure 2A**).

We posited that the foreground gene list length may have exacerbate the problem, so we performed
parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
A clear trend was observed for GO, Reactome, Cellmarkers and TFT GTRD, with Jaccard similarity
being significantly reduced when foreground gene lists were shorter than 500 genes.
Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
sets in the library.
Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.

![**Figure 2A. Effect of correcting the FDR problem.** Figures represent the mean number of
significant sets with and without correction of the FDR problem over seven independent RNA-seq
datasets.](../analysis/fig2_ORAfdr.png "Figure 2A shows that correcting the FDR bug slightly
reduces the number of significant sets.")

@@

Simulated differential expression profiles with apriori changes to predetermined pathway genes were
analysed in parallel by clusterProfilers enricher function, which suffers from the two aforementioned
problems and fora which does not.
In parallel, fgsea was used to compare the performance of these two ORA methods against a prototypical
FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.68 with added
variance of 1.0.
In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.9.
The recall of ORA methods dropped suddenly when added variance was 0.5 or larger.
fgsea showed stable precision of 0.94 across all levels of added variance, and superior recall overall.

@@

Here we have described the effect of two subtle problems with common implementations of ORA.
This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
obtained from clusterProfiler when using a custom set of pathways.
Our hunch that clusterProfiler was behaving in an undesired way was confirmed by several posts to
online bioinformatics forums dating to 2020 where confused users were wondering why the number of
background genes changed so much.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from the
background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described in the
official documentation.

The second problem is potentially more serious in that it could lead to lots of false positives,
as clusterProfiler underestimates the actual number of tests that were conducted.
As shown in Figure 2B, this affects results differently depending on the gene set library used and also
the number of genes selected in the foreground list.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).

@@

In addition to the subtle problems with ORA, there are some major ones too:

* Rarely used or reported properly (BG,FDR)

* Results vary dramatically depending on the number of genes selected in the foreground

* Less accurate as compared to FCS

The reason why such problems proliferated for so long is that using a custom background for ORA seems
to be the exception, rather than the rule.
Several tools do not even allow for the use of a custom background, suggesting that some tool developers
were unaware of the seriousness of this error.
This is one reason why ORA tool documentation also seems to lack any explanation of the importance of
using a custom background set.

>>>>

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets (FDR<0.05) over seven independent datasets detected with
clusterProfiler's `enricher` function with and without correcting the background problem.
Separate tests for up- and down-regulated gene lists were conducted.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

@@

A workaround for the FDR problem was developed, then original and corrected
analyses were conducted for the seven datasets and nine gene set libraries.
Results showed no major differences in the number of significant sets between
original and corrected analyses (**Figure 2A**).

We posited that the foreground gene list length might exacerbate the problem, so we performed
parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
A clear trend was observed for GO, Reactome, Cellmarkers and TFT GTRD, with Jaccard similarity
being significantly reduced when foreground gene lists were shorter than 500 genes.
Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
sets in the library.
Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.

![**Figure 2A. Effect of correcting the FDR problem.** Figures represent the mean number of
significant sets with and without correction of the FDR problem over seven independent RNA-seq
datasets. Foreground list consisted of 2000 genes with the smallest p-values](../analysis/fig2_ORAfdr.png
"Figure 2A shows that correcting the FDR bug slightly reduces the number of significant sets.")

@@

Simulated differential expression profiles with apriori changes to predetermined pathway genes were
analysed in parallel by clusterProfiler's `enricher` function, which suffers from the two
aforementioned problems and `fora` which does not.
In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
prototypical FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.61 with added
variance of 0.6.
In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.5.
The recall of ORA methods dropped suddenly when added variance was 0.2 or larger.
fgsea showed stable precision of 0.94 when added variance was 0.4 or lower, and dropped slightly to
0.91 with added variance of 0.6.
fgsea recall was better than either ORA tool across all levels of added variance.
This result shows that these two subtle problems result in poorer precision for clusterProfiler,
which becomes more severe with higher levels of noise.
Overall, fgsea had similar precision to fora (mean=0.94), but the higher recall of fgsea meant it had
a higher overall accuracy measured by F1 score (0.83) as compared to fora and clusterProfiler
(0.61 and 0.64 respectively).

@@

Here we have described the effect of two subtle problems with common implementations of ORA.
This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
obtained from clusterProfiler when using a custom set of pathways.
Our hunch that clusterProfiler was behaving in an undesired way was confirmed by the discovery of
several posts to online bioinformatics forums dating to ~2020 where confused users were wondering why
the number of background genes changed so much.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from
the background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described
in the official documentation.

The second problem is potentially more serious in that it could lead large numbers false positives
under specific conditions; when the foreground list is short and the median pathway size is relatively
small.
These false positives arise due to an underestimate of the actual number of tests conducted.
ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).

@@

Although the purpose of this work was to describe two subtle problems with ORA, it has some major
drawbacks which limit its utility.

* The method is rarely reported comprehensively, especially crucial details about background gene list
and FDR correction.

* Results vary dramatically depending on the number of genes selected in the foreground.

* Overall ORA is less accurate as compared to FCS

These bigger problems suggest FCS should replace ORA in most cases.
Where it makes sense to use ORA would be when FCS is not applicable, for example when investigating
the over-represented categories in gene expression modules defined by packages like WGCNA (Weighted
Gene Correlation Network Analysis)[@Langfelder2008-hu].
